A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-28 paint-contour-tree.tex

As part of the initialization, we determine
all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
vertices.

{\bf Painting:} The central tool is a notion 
of \emph{painting} of $\MM$. Initially associate a color with each maximum. Imagine there being a large
can of paint of a distinct color at each maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so every edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)

An \emph{initial} painting has the additional property that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
\end{definition}
%
Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
using a descending BFS from each maximum that does not explore previously colored edges. 
As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.

{\bf Colors of vertices:} Note that a painting only colors edges, not vertices. We associate certain
sets of colors with vertices.

\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
A vertex $v$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
edge with $v$ at the lower endpoint. For $v$, $\col(v)$ is the set of colors that touch $v$.
\end{definition}

Abusing notation, we say a contour $\phi$ is of color $c$ if all edges that intersect $\phi$ have color $c$.

\begin{definition} \label{def:color-full} Consider vertex $v$. An \emph{up-star} of $v$ is the set 
of edges that all connect to a fixed component of $\MM^+_v$. (Each $v$
participates in at mot $2$ up-stars.) A color $c \in \col(v)$ \emph{fully touches} $v$
if all edges in an up-star are colored $v$.
%and let $\phi$ be a $(f(v)+\delta)$-contour in
%$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
%that fully touch $v$ is denoted $\fcol(v)$.
\end{definition}

%Note that $\fcol(v)$ has size at most $2$, by the Morse condition. 
Note that $\col(v)$ 
%and $\fcol(v)$ can
can be initially determined in $O(|\MM|)$ time, by a simple linear scan.

\begin{proof} Since $e$ has color $c$,
there must exist vertices in $\MM^+_v$ touched by $c$. Consider the highest
vertex $w$ in $\MM^+_v$ that is touched by $c$ and some other color. If no such vertex exists,
this means all edges incident to a vertex touched by $c$ are colored $c$. By walking through
$\MM^+_v$, we deduce that all edges are colored $c$.

So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence $c$ fully touches $w$. 
But $w$ is touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
must have up-degree $2$ and is critical.
\end{proof}

@@

%While this is the intuitive viewpoint, we describe how this information is actually maintained.
The vertices $\{\h(\rep(c))\}$ represent the set of contours $\Phi$. In terms of the contour tree, 
we have already constructed the contour trees rooted at each $h(\rep(c))$. This is trivially true at the initialization,
when each $h(c)$ is a maximum. One can now treat each $h(\rep(c))$ as a maximum, corresponding to (a possibly new) color $\rep(c)$.
To reduce confusion between the original painting and the new colors, we will assume that $\col(v)$ and $\fcol(v)$
are as given by the original painting.

@@

>>>>

%
%As part of the initialization, we determine
%all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
%in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
%of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
%vertices.

{\bf Painting:} The central tool is a notion 
of \emph{painting} $\MM$. Initially associate a color with each maximum. Imagine there being a large
can of paint of a distinct color at each maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so every edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$.

An \emph{initial} painting has the additional property that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
\end{definition}
%
%Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
%using a descending BFS from each maximum that does not explore previously colored edges. 
%As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.

A painting only colors edges, not vertices. We associate certain
sets of colors with vertices.

\begin{definition} \label{def:color-set} Fix a painting $\chi$ and vertex $v$.
\begin{asparaitem}
	 \item An \emph{up-star} of $v$ is the set of edges that all connected to a fixed component of $\MM^+_v$.
	 \item A vertex $v$ is \emph{touched by color $c$} if $x$ is incident to a $c$-colored
	 edge with $v$ at the lower endpoint. For $v$, $\col(v)$ is the set of colors that touch $v$.
	 \item A color $c \in \col(v)$ \emph{fully touches} $v$ if all edges in an up-star are colored $v$.
\end{asparaitem}
\end{definition}

Each non-maxima $v$ participates in at least $1$ and at most $2$ up-stars (the latter iff
$v$ is a join).

\ignore{
Abusing notation, we say a contour $\phi$ is of color $c$ if all edges that intersect $\phi$ have color $c$.

\begin{definition} \label{def:color-full} Consider vertex $v$. An \emph{up-star} of $v$ is the set 
of edges that all connected to a fixed component of $\MM^+_v$. 
%and let $\phi$ be a $(f(v)+\delta)$-contour in
%$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
%that fully touch $v$ is denoted $\fcol(v)$.
\end{definition}

%Note that $\fcol(v)$ has size at most $2$, by the Morse condition. 
Note that $\col(v)$ 
%and $\fcol(v)$ can
be initially determined in $O(|\MM|)$ time, by a simple linear scan.
}

@@

%While this is the intuitive viewpoint, we describe how this information is actually maintained.
The vertices $\{\h(\rep(c))\}$ represent the set of contours $\Phi$. In terms of the contour tree, 
we have already constructed the contour trees rooted at each $h(\rep(c))$. This is trivially true at the initialization,
when each $h(c)$ is a maximum. One can now treat each $h(\rep(c))$ as a maximum, corresponding to (a possibly new) color $\rep(c)$.
To reduce confusion between the original painting and the new colors, we will assume that $\col(v)$ and $\fcol(v)$
are as given by the original painting. 
}

@@

Our main workhorse is the following technical lemma.

\begin{proof} Since $e$ has color $c$,
there must exist vertices in $\MM^+_v$ touched by $c$. Consider the highest
vertex $w$ in $\MM^+_v$ that is touched by $c$ and some other color. If no such vertex exists,
this means all edges incident the a vertex touched by $c$ are colored $c$. By walking through
$\MM^+_v$, we deduce that all edges are colored $c$.

So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence, $c$ fully touches $w$. 
But $w$ is touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
must have up-degree $2$ and is critical.
\end{proof}

We prove a series of claims that will lead to the correctness proof.

\begin{claim} \label{clm:upstar} Consider ripe vertex $v$ and take the up-star connecting
to some component of $\MM^+_v$. All edges in this component and the up-star have the same color.
\end{claim}

\begin{proof} Let $c$ be the color of some edge in this up-star
By ripeness, $v$ is the highest in $T(\rep(c))$.
Denote the component of $\MM^+_v$ by $\PP$.
By \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or there exists critical vertex $w \in \PP$
fully touched by $\rep(c)$ and another color. In the latter case, $w$ has not been processed,
so $w \in T(\rep(c))$ (contradiction to ripeness). Therefore, all edges in $\PP$ are colored $\rep(c)$.
\end{proof}

\begin{claim} \label{clm:process} The partial output on the processed vertices is exactly
the restriction of $\jc(\MM)$ to these vertices.
\end{claim}

\begin{proof} More generally, we prove the following: all outputs on processed vertices
are edges of $\cC(\MM)$ and for any color $c$, $\h(c)$ is the lowest processed vertex
of that color. We prove by induction on the processing order. Initially, this is trivially
true, handling these base case. For the induction step, consider the situation
when $v$ is being processed.

The vertex $v$ is ripe. Take any up-star, and the corresponding component $\PP$
of $\MM^+_v$ that it connects to. By \Clm{upstar}, all edges in $\PP$ and the up-star
have the same color (say $c$). If some critical vertex in $\PP$ is not processed,
it must be in $T(c)$, which violates the ripeness of $v$.
Thus, all critical vertices in $\PP$ have been processed.
